Type-Safe Modular Parsing

### 2016-12-14 paper/sections/Overview.tex
Done 2.1-2.3

Modular parsing based on parser generators is supported by many libraries []. Users can separate the concrete/abstract syntax and related parsing code into reusable components. Then the corresponding parsers are built by their library utilities. For example, NOA [] uses Java annotation processing to collect grammar information, and then generates ANTLR4 parsers. However, such generation procedure requires a whole compilation after the collection of all grammar pieces. Once the grammar changed, even slightly, grammar information must be re-collected and the parser must be re-generated. Hence thoes libraries only have syntactic modularity.

@@

Unfortunatelly many parser combinators have important limitations for
modular parsing. In particular several parser combinators,
including the famous library Parsec~\cite{} library, require
programmers to manually do \textit{left-recursion elimination}, and
require significant amounts \textit{backtracking}. Both of these are
problematic in a modular setting.

\paragraph{Left-Recursion Elimination} The main problem with
left-recursion elimination is that it is a \emph{global}
transformation on a grammar. Given that the full grammar is known,
then it is possible to remove all left-recursive cases.
\bruno{Illustrate with a concrete example: show a simple
  left-recursive grammar; then do left-recursion elimination. Then add
a new case into the original left-recursive grammar; and show that
a very different grammar is obtained after left-recursion elimination.}

However, when doing modular parsing, the full grammar is not known!

Parser combinators adopt top-down, recursive descent parsing, which
uses backtracking to search through the possible branches. However,
simple backtracking parser combinators such as the famous library
Parsec~\cite{} in Haskell, have some drawbacks. One is that they cannot
support left-recursive grammars. The common solution is to rewrite a
left-recursive grammar into an equivalent one, so called
\textit{left-recursion elimination}. This solution requires
information of the whole grammar, so that it cannot be applied in a
modular setting. Because the grammar may be extended in the future,
essentially we only have parts of the grammar.

@@

\paragraph{Packrat Parsing}
Fortunatelly more advanced parsing techniques such as Packrat
parsing~\cite{}, address the limitations of simple parser combinators
such as Parsec. Packrat parsers use
memorization to record the result of applying each parser at each
position of the input, so that repeated computation is eliminated, and
they supports left-recursive grammars directly. All of these properties are very
suitable for modularity, thus we use Packrat parsers as the underlying
parsing technique. In section \ref{sec:packratparsers}, we will give
an overview of the Packrat parsing library implemented in Scala,
which we use.

@@

\huang{partly done, only extend a new language construct in the example, how about leaving extension of new operations in the OA subsection?}\bruno{I think we need to set up a challenge here, similar to the challange of
  the expression problem. The challange should be like. Build a parser
for a simple expression language, then extend the language with both
a new language construct and a new operation. This section will show
how we can do that using traditional parsers, but without modular
type-safety and separate compilation. The remaining sections will show
that the techniques introduced by us, enable us to solve those two challenges.}

@@

\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <var>
    \alt <expr> <expr>
    \alt `(' <expr> `)'
\end{grammar}

@@

\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= ...
    \alt `\\' <var> `.' <expr>
\end{grammar}

>>>>

Modular parsing based on parser generators is supported by many libraries []. Users can separate the concrete/abstract syntax and related parsing code into reusable components. Then the corresponding parsers are built by their library utilities. For example, NOA [] uses Java annotation processing to collect grammar information, and then generates ANTLR4 parsers. However, such generation procedure requires a whole compilation after the collection of all grammar pieces. Once the grammar changed, even slightly, grammar information must be re-collected and the parser must be re-generated. Hence those libraries only have syntactic modularity.

@@

Unfortunately many parser combinators have important limitations for
modular parsing. In particular several parser combinators,
including the famous library Parsec~\cite{} library, require
programmers to manually do \textit{left-recursion elimination}, and
require significant amounts \textit{backtracking}. Both of these are
problematic in a modular setting.

\paragraph{Left-Recursion Elimination} The top-down, recursive descent parsing strategy adopted by those parser combinator libraries cannot support left-recursive grammars directly. The common solution is to rewrite the grammar into an equivalent but not left-recursive one, so called left-recursion elimination.

A left-recursive grammar of adding integers can be rewritten as below. On the right side, we got a new grammar without left-recursion.

\begin{tabular}{m{0.4\linewidth}m{0.4\linewidth}}
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int> \alt <expr> `+' <int>
\end{grammar}
&
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int> <expr'>

<expr'> ::= <empty> \alt `+' <int> <expr'>
\end{grammar}
\end{tabular}

The main problem with left-recursion elimination is that it is a \emph{global} transformation on a grammar. Given that the full grammar is known, then it is possible to remove all left-recursive cases. However, when doing modular parsing, the full grammar is not known!

For the example above, if we extend the original grammar to support subtraction, we must analyse the full grammar again to rewrite it.

\begin{tabular}{m{0.4\linewidth}m{0.4\linewidth}}
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int> \alt <expr> `+' <int> \alt <expr> `-' <int>
\end{grammar}
&
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int> <expr'>

<expr'> ::= <empty> \alt `+' <int> <expr'> \alt `-' <int> <expr'>
\end{grammar}
\end{tabular}

Another issue of left-recursion elimination is that it requires extra bookkeeping work to retain the original semantic. For example, the expression $1+2-3$ is parsed as $(1+2)-3$ in the left-recursive grammar, but after rewrite the result is $1((+2)-3))$. The parse tree must be transformed to recover its structure.

\huang{done, the example may be not very good, but i haven't found a better one}\bruno{Illustrate with a concrete example: show a simple
  left-recursive grammar; then do left-recursion elimination. Then add
a new case into the original left-recursive grammar; and show that
a very different grammar is obtained after left-recursion elimination.}

@@

\paragraph{Packrat Parsing}
Fortunately more advanced parsing techniques such as Packrat
parsing~\cite{}, address the limitations of simple parser combinators
such as Parsec. Packrat parsers use
memorization to record the result of applying each parser at each
position of the input, so that repeated computation is eliminated, and
they supports left-recursive grammars directly. All of these properties are very
suitable for modularity, thus we use Packrat parsers as the underlying
parsing technique. In section \ref{sec:packratparsers}, we will give
an overview of the Packrat parsing library implemented in Scala,
which we use.

@@

\huang{partly done, only extend a new language construct in the example, how about leaving extension of new operations in the OA subsection?}\bruno{I think we need to set up a challenge here, similar to the challenge of
  the expression problem. The challenge should be like. Build a parser
for a simple expression language, then extend the language with both
a new language construct and a new operation. This section will show
how we can do that using traditional parsers, but without modular
type-safety and separate compilation. The remaining sections will show
that the techniques introduced by us, enable us to solve those two challenges.}

@@

\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <ident>
    \alt <expr> <expr>
    \alt `(' <expr> `)'
\end{grammar}

@@

\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= ...
    \alt `\\' <ident> `.' <expr>
\end{grammar}
